\chapter{Installation instructions}
\label{attachment:setup}

To set up and run the application, follow these instructions.

The code is available online. Front end\footnote{\url{https://github.com/kszabova/rhythmify_frontend/tree/bc_final}}
and back end\footnote{\url{https://github.com/kszabova/rhythmify_backend/tree/bc_final}} are available separately.
Download both repositories and store them on your device.

Install the correct version of MAFFT for your platform.\footnote{\url{https://mafft.cbrc.jp/alignment/software/}} If you
are using Windows, choose the option for installing it on WSL.\footnote{\url{https://docs.microsoft.com/en-us/windows/wsl/install-win10}}
In this case, please, ensure that MAFFT can be accessed by running \verb|wsl mafft <options>| from command line.

Ensure that you have both Python (\textgreater =3.9.2) and Node.js installed. If you do not, follow the installation instructions
on their respective web pages\footnote{\url{https://www.python.org/downloads/}}\footnote{\url{https://nodejs.org/en/download/}}

Set up the Django framework by following the installation instructions.\footnote{\url{https://www.djangoproject.com/download/}}
Do the same for the Angular platform.\footnote{\url{https://angular.io/guide/setup-local}}

Once all these dependencies are installed, navigate to the back end directory. In command line, run \verb|pip install -r requirements.txt|.
Additionally, it is necessary to download the Latin corpus for the correct processing of Latin texts.\footnote{\url{https://docs.cltk.org/en/latest/data.html}}
In Python environment, run the following commands:

\begin{verbatim}
    from cltk.data.fetch import FetchCorpus
    corpus_downloader = FetchCorpus(language="lat")
    corpus_downloader.import_corpus("lat_models_cltk")
\end{verbatim}

To run the back end, type \verb|python manage.py runserver|.

Now, navigate to the front end directory and run \verb|npm install|. This will install the required packages.
To run the front end, type \verb|ng serve|.
The application can be accessed from your browser on the address \verb|localhost:4200|.
